Local Spectral Clustering of Density Upper Level Sets

### 2019-11-03 arxiv/local_spectral_density_clustering_arxiv.tex
alden arxiv draft complete

%%% Graph quantities
\newcommand{\Cest}{\widehat{C}}
\newcommand{\degminpr}{\deg_{\min}'}
\newcommand{\degminwt}{\widetilde{\deg}_{\min}}
\newcommand{\degmaxwt}{\widetilde{\deg}_{\max}}
\newcommand{\degmax}{\deg_{\max}}
\newcommand{\piminwt}{\widetilde{\pi}_{\min}}
\newcommand{\piminpr}{\pibf_{\min}'}
\newcommand{\degmin}{\deg_{\min}}

@@

\begin{abstract}
\noindent We analyze the Personalized PageRank (PPR) algorithm, a local spectral method for clustering,
which extracts clusters using locally-biased random walks around a user-specified
seed node.  In contrast to previous work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also provide empirical support for our theory.
\end{abstract}
\end{center}
\section{Introduction}
\label{sec: introduction}
In this paper, we consider the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  Our particular focus is on
spectral clustering methods, a family of powerful
nonparametric clustering algorithms. Roughly speaking, a spectral algorithm
first constructs a geometric graph $G$, where vertices correspond to samples,
and edges correspond to proximities between samples. The algorithm 
then estimates a feature
embedding based on (an appropriate) Laplacian matrix of the graph 
$G$, and applies a simple clustering
technique (like $k$-means clustering) in the embedded feature space.

@@

\subsection{PPR on a neighborhood graph} 
We now describe the clustering
algorithm that will be our focus for the rest of the paper. Let $\Xbf = \{x_1,
\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$,
with density~$f$.  For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the
\emph{$r$-neighborhood graph} of $\Xbf$, an unweighted, undirected graph with
vertices $V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i -
x_j} \leq r$, where $\norm{\cdot}$ is the $\ell_2$ norm. We denote by $\Abf \in
\Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if
$(u,v) \in E$ and $0$ otherwise.  We also denote by $\Dbf$ the diagonal degree
matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n
\times n$ identity matrix.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, based on
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be 
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in the
$v$-th position and 0 elsewhere). For a level $\beta > 0$ and a target volume
$\vol_0 > 0$, we define a \emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$
as   
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta := \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$}\footnote{The choice of a specific range such as  
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} that has minimum normalized cut.
%\smash{$\Phi(S_{\beta^*}; G_{n,r})$}. 
For concreteness, the PPR algorithm is summarized in Algorithm~\ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary volume
$\vol_0 > 0$. \\     
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $p=\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn: 
    ppr_vector}. 
  \STATE For \smash{$\beta \in (\frac{1}{40}, \frac{1}{11})$} compute sweep cuts 
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return the cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (\frac{1}{40}, \frac{1}{11})}~ \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Estimation of density clusters} 
Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The
size of the symmetric set difference between estimated and empirical clusters is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}.

\begin{definition}[Symmetric set difference]
  \label{def: symmetric_set_diff}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and set
  $\mathcal{S} \subseteq \Reals^d$, we define   
  \begin{equation}
    \label{eqn: misclassification_rate}
    \Delta(\Cest, \mathcal{S}) := \abs{\Cest \setminus \mathcal{S}[\Xbf] \cup
      \mathcal{S}[\Xbf] \setminus \Cest},
  \end{equation}
  the cardinality of the symmetric set difference between 
  \smash{$\Cest$} and $\mathcal{S} \cap \Xbf = \mathcal{S}[\Xbf]$. 
\end{definition}

However, the symmetric set difference does not measure whether \smash{$\Cest$} 
can distinguish any two distinct clusters \smash{$\Cset,\Cset' \in
  \Cbb_f(\lambda)$}. We therefore also study a second notion of cluster
estimation, first introduced by \citet{hartigan1981}, and defined
asymptotically.

\begin{definition}[Consistent density cluster estimation]
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
  \smash{$\Cset \in \Cbb_f(\lambda)$}, we say \smash{$\Cest$} is a consistent 
  estimator of $\Cset$ if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with
  $\Cset \not= \Cset'$, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}
\noindent Consistent cluster recovery roughly ensures that, for a given $\lambda$, the estimated cluster $\Cest$ contains all
points in a true
density cluster $\Cset$, and simultaneously does not contain any points in any other density cluster $\Cset'$.

@@

\subsection{Related Work and Our Contributions}
In addition to the background on local spectral
clustering given previously, a few related lines of work are worth
highlighting. \citet{shi2009} and \citet{schiebinger2015} examine the consistency of  
spectral algorithms in recovering the latent labels in certain
nonparametric mixture models. Their results focus on global rather than local 
methods, and thus impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density 
clusters, which is the focus in our work. Perhaps most importantly, these works
rely on general cluster saliency conditions, which implicitly depend on many
distinct geometric aspects of the cluster $\Cset$ under consideration. We make
this dependence more explicit, and in doing so expose the role each geometric
condition plays in the clustering problem.

More broadly, density clustering and level set estimation is a well-studied
problem. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level-set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010, balakrishnan2013} consider consistent estimation of the
cluster tree.
%, to note but a few works. 
We emphasize that our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is to start with a 
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

A summary of our results (and an outline for this
paper) is as follows.

@@

@@

\noindent At this point we summarize one of our main takeaways: 
PPR, run on a neighborhood graph, recovers \emph{geometrically
compact} high-density clusters. Our main theoretical results, make this takeaway 
precise, and provide concrete ways of quantifying the geometric
compactness of a density cluster. To complement this, we also provide experimental
evidence that the PPR algorithm can fail to recover high-density clusters which are not
geometrically compact.

\section{Main Results}
\label{sec: consistent_cluster_estimation_with_ppr}
In this section we present our main results on the accuracy of the PPR algorithm 
for recovering density clusters. 
We begin by formally introducing a variety of geometric conditions, 
and use these to define a condition
number $\kappa(\Cset)$, which measures the difficulty PPR will have in  
estimating a density cluster $\Cset$.
With this condition in place 
our first main result (Theorem~\ref{thm: misclassification_rate}) provides a bound
on the symmetric set difference between the estimated cluster $\Cest$,
obtained by an appropriately initialized version of the PPR algorithm,
and
the target cluster $\Cset$.
Our next main result (Theorem~\ref{thm: consistent_recovery_of_density_clusters})
shows that for sufficiently well-conditioned target clusters $\Cset$, the output 
of the PPR algorithm $\Cest$
is consistent in the sense defined in Definition~\ref{def: consistent_density_cluster_estimation}.

@@

\item
  \label{asmp: cluster_separation}
  \emph{Cluster separation:}
  For all clusters $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$, 
  $\dist(\Csig,\Csig') > \sigma$, where $\dist(\Csig,\Csig') := \inf_{x
    \in \Csig} \dist(x,\Csig')$.

\item 
  \label{asmp: low_noise_density}
  \emph{Low noise density:} There exist $\gamma,c_0 > 0$ such that for 
  $x \in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
  \begin{align*}
  \inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma}. 
  \end{align*}
  Roughly, 
  this assumption ensures that the density decays sufficiently quickly as we move away from
  the target cluster $\Csig$, and is a standard assumption in the level-set estimation literature
  (see for instance~\cite{singh2009}).

  %%% AJG 5/20: Should I turn this into two conditions?
\item
  \label{asmp: embedding}
  \emph{Lipschitz embedding:}
  There exists $g: \Reals^d \to \Reals^d$ with the following properties: 
  \begin{enumerate}
  \item we have $\Csig = g(\mathcal{K})$, for a convex set $\mathcal{K} \subseteq \Rd$
  with $\mathrm{diam}(\mathcal{K}) = \sup_{x,y \in \mathcal{K}}\norm{x - y} =:
  \diam < \infty$
  \item $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, where
  $\nabla g(x)$ is the Jacobian of $g$ evaluated at $x$ and
  \item for some $L
  \geq 1$,   
  \begin{equation*}
    \frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~
    \text{for all $x,y \in \mathcal{K}$}. 
  \end{equation*}
  \end{enumerate}
  Succinctly, we assume that $\Csig$ is the image of a convex set with finite diameter 
  under a  measure preserving, bi-Lipschitz transformation.

\item
  \label{asmp: bounded_volume}
  \emph{Bounded volume:}
  We assume that neighborhood graph radius $0 < r \leq \sigma/2d$ is chosen such that
  \begin{equation*}
    2 \int_{\Csig} \Pbb(B(x,r)) f(x) dx \leq \int_{\Rd} \Pbb(B(x,r)) f(x) dx,
  \end{equation*}
  where $B(x,r)$ is the closed ball of radius $r$ at $x$.
\end{enumerate}
To motivate these conditions, we provide a brief high-level sketch
of our analysis (we return to this more formally in Section~\ref{sec: analysis}).
\citet{zhu2013} show for an arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is  the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}, and our main goal will be to upper bound 
  this quantity. 
  Towards this goal, as we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. In more detail, assumptions  
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound on diameter in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.

>>>>

%%% Graph quantities
\newcommand{\Cest}{\widehat{C}}
\newcommand{\degminpr}{d_{\min}'}
\newcommand{\degminwt}{\widetilde{d}_{\min}}
\newcommand{\degmaxwt}{\widetilde{d}_{\max}}
\newcommand{\degmax}{d_{\max}}
\newcommand{\piminwt}{\widetilde{\pi}_{\min}}
\newcommand{\piminpr}{\pibf_{\min}'}
\newcommand{\degmin}{d_{\min}}

@@

\begin{abstract}
\noindent We analyze the Personalized PageRank (PPR) algorithm, a local spectral method for clustering,
which extracts clusters using locally-biased random walks around a user-specified
seed node.  In contrast to previous work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters, and conversely fails to recover geometrically poorly conditioned density clusters. We also provide empirical support to illustrate our theoretical results.
\end{abstract}
\end{center}
\section{Introduction}
\label{sec: introduction}
In this paper, we consider the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  Our particular focus is on
spectral clustering methods, a family of powerful
nonparametric clustering algorithms. Roughly speaking, a spectral algorithm
first constructs a geometric graph $G$, where vertices correspond to samples,
and edges correspond to proximities between samples. The algorithm 
then estimates a feature
embedding based on (an appropriate) Laplacian matrix of the graph 
$G$, and applies a simple clustering
technique (like $k$-means clustering) in the embedded feature space.

@@

\subsection{PPR on a neighborhood graph} 
We now describe the clustering
algorithm that will be our focus for the rest of the paper. Let $\Xbf = \{x_1,
\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$,
with density~$f$.  For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the
\emph{$r$-neighborhood graph} of $\Xbf$, an unweighted, undirected graph with
vertices $V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i -
x_j} \leq r$, where $\norm{\cdot}$ is the Euclidean norm. We denote by $\Abf \in
\Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if
$(u,v) \in E$ and $0$ otherwise.  We also denote by $\Dbf$ the diagonal degree
matrix, with entries $\Dbf_{uu} := \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n
\times n$ identity matrix.

Next, we define the PPR vector $\pbf_v = \pbf(v,\alpha;G_{n,r})$, based on
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be 
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf_v = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in the
$v$-th position and 0 elsewhere).

\subsection{Sweep cut.}
\label{subsection: sweep_cut}

We employ the sweep cut method to obtain a local cluster from the PPR embedding $\pbf_v$. For a level $\beta > 0$, we define a \emph{$\beta$-sweep cut} of $\pbf_v = (\pbf_v(u))_{u \in V}$
as   
\begin{equation}
\label{eqn: sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{\pbf_v(u)}{\Dbf_{uu}} > \beta}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(L, U)$} (where the range $(L,U)$ is also a user-specified parameter), we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} that has minimum normalized cut.
%\smash{$\Phi(S_{\beta^*}; G_{n,r})$}. 
For concreteness, the PPR algorithm is summarized in Algorithm~\ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, sweep cut range $(L,U)$. \\     
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf_v=\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn: 
    ppr_vector}. 
  \STATE For \smash{$\beta \in (L,U)$} compute sweep cuts 
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return the cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (L,U)}~ \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Estimation of density clusters} 
Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The size of the symmetric set difference between estimated and empirical cluster is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}. We will consider a related metric, the volume of the symmetric set difference, which weights points according to their degree in $G_{n,r}$. 
\begin{definition}
	\label{def:volume_symmetric_set_difference}
	For an estimator $\widehat{C} \subseteq \Xbf$ and a set $\mathcal{S} \subseteq \Rd$, the symmetric set difference is 
	\begin{equation*}
	\widehat{C} \vartriangle \mathcal{S}[\Xbf] := \bigl(\widehat{C} \cap \setminus \mathcal{S}[\Xbf]\bigr) \cup \bigl(\Xbf\setminus\widehat{C} \cap \mathcal{S}[\Xbf]\bigr).
	\end{equation*}
	We will measure the discrepancy of $\widehat{C}$ to the set $\mathcal{S}[\Xbf]$ by the volume of the symmetric set difference,
	\begin{equation*}
	\Delta(\widehat{C}, \mathcal{S}[\Xbf]) := \vol(\widehat{C} \vartriangle \mathcal{S}[\Xbf]; G_{n,r})
	\end{equation*}
\end{definition}
However, the symmetric set difference does not measure whether \smash{$\Cest$} 
can distinguish any two distinct clusters \smash{$\Cset,\Cset' \in
\Cbb_f(\lambda)$}. We therefore also study a second notion of cluster
estimation, first introduced by \citet{hartigan1981}, and defined
asymptotically.

\begin{definition}[Consistent density cluster estimation]
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
  \smash{$\Cset \in \Cbb_f(\lambda)$}, we say \smash{$\Cest$} is a consistent 
  estimator of $\Cset$ if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with
  $\Cset \not= \Cset'$, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}
\noindent Consistent cluster recovery roughly ensures that, for a given density threshold $\lambda$, the estimated cluster $\Cest$ contains all
points in a true
density cluster $\Cset \in \Cbb_f(\lambda)$, and simultaneously does not contain any points in any other density cluster $\Cset' \in \Cbb_f(\lambda)$.

@@

\subsection{Related Work and Our Contributions}
In addition to the background on local spectral
clustering given previously, a few related lines of work are worth
highlighting. In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. The ability of spectral algorithms to perform clustering -- community detection-- in this context is well-understood. In particular, \citet{rohe2011} upper bound the fraction of nodes misclassified by a spectral algorithm for the high-dimensional (large number of blocks) SBM,  \citet{lei2015} extend these results to the sparse (low average degree) regime, and \citet{balakrishnan2011} analyze the misclassification rate when the block model exhibits some hierarchical structure. The framework we consider, where nodes correspond to data points sampled from an underlying density, and edges between nodes are formed based on geometric proximity, is quite different than the SBM, and therefore so is our analysis.

The study of spectral algorithms on neighborhood graphs has focused on establishing consistency results-- that is, asymptotic convergence of eigenvalues and eigenvectors of matrices to eigenvalues and eigenfunctions of limiting operators. \citet{koltchinskii2000} establish convergence of spectral projections of the adjacency matrix to a limiting integral operator, with similar results obtained using simplified proofs in \citet{rosasco10}. \citet{vonluxburg2008} studies convergence of eigenvectors of the Laplacian matrix for a neighborhood of fixed radius. \citet{belkin07} and \citet{garciatrillos18} extend these results to the regime where the radius $r \to 0$ as $n \to \infty$.

These results are of fundamental importance; however, the behavior of the spectra of these continuum operators can in general be hard to grasp. Therefore, further work relating this spectra to the geometry of the distribution $\Pbb$ is of interest. In this spirit, \citet{shi2009,schiebinger2015,garciatrillos19} examine the ability of  
spectral algorithms to recover the latent labels in certain geometrically well-conditioned nonparametric mixture models. Their results focus on global rather than local 
methods, and thus impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density 
clusters, which is the focus in our work. Perhaps most importantly, these works
rely on general cluster saliency conditions, which implicitly depend on many
distinct geometric aspects of the cluster $\Cset$ under consideration. We make
this dependence more explicit, and in doing so expose the role each geometric
condition plays in the clustering problem.

A substantial portion of our technical effort involves estimating graph-theoretic connectivity properties of empirical density clusters in the neighborhood graph $G_{n,r}$. In particular, we prove an upper bound on the mixing time of a random walk run only over the subset of nodes in $G_{n,r}$ which fall within a density cluster $\Cset$. To do so, we rely on a series of seminal works upper bounding the mixing time of \emph{geometric random walks} (see \citet{vempala2005} for a comprehensive review.)  This study was initiated by \citet{dyer1991}, who used geometric random walks as a fundamental subroutine to efficiently compute volumes of high-dimensional convex bodies. These results are improved in \citet{lovasz1990,kannan97,kannan06}, who show \textit{inter alia} that the bounds on mixing time can be sharpened by avoiding so-called ``start-penalties''.(These improvements are crucial to our work, for they allow us to avoid factors of $\log n$ which would imply divergence as $n \to \infty$.) Following the lead of \citet{abbasi-yadkori2016,abbasi-yadkori2016a}, we extend these results to hold for Lipschitz deformations of convex sets. Additionally, we relate the mixing time of these (continuous-space) geometric random walks to the mixing time of random walks over (discrete) neighborhood graphs.

Finally, it is worth mentioning that density clustering and level set estimation are well-studied problems. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level-set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010, balakrishnan2013} consider consistent estimation of the
cluster tree.
%, to note but a few works. 
We emphasize that our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is to start with a 
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

Specifically, a summary of our results (and an outline for this
paper) is as follows.

@@

\item In Section \ref{sec:lower_bound}, we provide an accompanying lower bound, demonstrating that when the cluster $\Cset$ is sufficiently poorly conditioned, it will not be recovered by Algorithm~\ref{alg: ppr}.

@@

\noindent At this point we summarize one of our main takeaways: 
PPR, run on a neighborhood graph, recovers only \emph{geometrically
compact} high-density clusters. Our main theoretical results, make this takeaway 
precise, and provide concrete ways of quantifying the geometric
compactness of a density cluster.

\section{Main Results}
\label{sec: consistent_cluster_estimation_with_ppr}
In this section we present our main results on the accuracy of the PPR algorithm 
for recovering density clusters. 
We begin by formally introducing a variety of geometric conditions, 
and use these to define a condition
number $\kappa(\Cset)$, which measures the difficulty PPR will have in  
estimating a density cluster $\Cset$.
With this condition in place 
our first main result (Theorem~\ref{thm:volume_ssd}) provides a bound
on the symmetric set difference between the estimated cluster $\Cest$,
obtained by an appropriately initialized version of the PPR algorithm,
and
the target cluster $\Cset$.
Our next main result (Theorem~\ref{thm: consistent_recovery_of_density_clusters})
shows that for sufficiently well-conditioned target clusters $\Cset$, the output 
of the PPR algorithm $\Cest$
is consistent in the sense defined in Definition~\ref{def: consistent_density_cluster_estimation}.

@@

\item 
  \label{asmp: low_noise_density}
  \emph{Low noise density:} There exists $c_0 > 0$ and $\gamma \in [0,1]$ such that for any $x \in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
  \begin{align*}
  \inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma}. 
  \end{align*}
  Roughly, 
  this assumption ensures that the density decays sufficiently quickly as we move away from
  the target cluster $\Csig$, and is a standard assumption in the level-set estimation literature
  (see for instance~\cite{singh2009}).

\item
  \label{asmp: embedding}
  \emph{Lipschitz embedding:}
  There exists $g: \Reals^d \to \Reals^d$ with the following properties: 
  \begin{enumerate}
  \item we have $\Csig = g(\mathcal{K})$, for a convex set $\mathcal{K} \subseteq \Rd$
  with $\mathrm{diam}(\mathcal{K}) = \sup_{x,y \in \mathcal{K}}\norm{x - y} =:
  \diam < \infty$
  \item $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, where
  $\nabla g(x)$ is the Jacobian of $g$ evaluated at $x$ and
  \item for some $L
  \geq 1$,   
  \begin{equation*}
    \norm{g(x) - g(y)} \leq L \norm{x - y} ~
    \text{for all $x,y \in \mathcal{K}$}. 
  \end{equation*}
  \end{enumerate}
  Succinctly, we assume that $\Csig$ is the image of a convex set with finite diameter 
  under a  measure preserving, Lipschitz transformation.

\item
  \label{asmp: bounded_volume}
  \emph{Bounded volume:}
  For a set $\mathcal{S} \subseteq \Reals^d$, define the $\Pbb$-weighted volume of $\mathcal{S}$ to be
  \begin{equation}
  \label{eqn:volume}
  \vol_{\Pbb,r}(\mathcal{S}) := \int_{\mathcal{S}} \Pbb(B(x,r)) \,d\Pbb(x).
  \end{equation}
  where $B(x,r)$ is the closed ball of radius $r$ centered at $x$.

  We assume that neighborhood graph radius $0 < r \leq \sigma/2d$ is chosen such that
  \begin{equation*}
  \vol_{\Pbb,r}(\mathcal{S}) \leq \frac{1}{2}\vol_{\Pbb,r}(\Reals^d).
  \end{equation*}

\end{enumerate}

To motivate these conditions, we provide a brief high-level sketch
of our analysis (we return to this more formally in Section~\ref{sec: analysis}). 
\citet{zhu2013} show for an arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \vartriangle S; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is  the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn: graph_symmetric_set_difference_1} is precisely
the volume of the symmetric set difference, and our main goal is therefore to upper bound the graph functionals $\Phi$ and $\tau_{\infty}$.
  Towards this goal, as we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. In more detail, assumption 
\ref{asmp: low_noise_density} yields an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density} and \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound on diameter in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.

\subsubsection{Condition number} 
Motivated by \eqref{eqn:
  graph_symmetric_set_difference_1}, we will define $\kappa(\Cset)$ to be an
upper bound on the product \smash{$\Phi(\Csig[\Xbf]; G_{n,r}) \cdot
  \tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. Following, the analysis of \citet{zhu2013}
  we see that the smaller $\kappa(\Cset)$ is, the
more success PPR will have in recovering the target cluster $\Cset$.
